---
title: "Per Individual Read Depth Tutorial"
author: "Quinn W"
date: "8/6/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(vcfR)

```

Here, we look at the distribution of the depth of the most abundant allele on a per individual basis. This can be helpful for developing filtering criteria.

First, we'll read in a vcf file. 


```{r, message = FALSE, results = 'hide'}

chilo_90 <- read.vcfR("./Data/chilo/all_filtered_chilo_90.vcf")

```

Here is the function to generate the distribution plots for each individual. The red lines correspond to the 15th and 95th percentiles.

```{r}

#takes vcfR object and plots read depth distribution for the most abundant allele
plot_read_depth <- function(vcf) {
  #extract allele depths
  ad <- extract.gt(vcf, element = 'AD')
  #get depth of the most abundant allele
  a1 <- masplit(ad, record = 1)
  tmp <- a1[,1]
  sums <- apply(a1, MARGIN=2, quantile, probs = c(.15,.95), na.rm=TRUE)


  for (i in 1:ncol(ad)) {
    tmp <- a1[,i]
    sums <- apply(a1, MARGIN=2, quantile, probs = c(.15,.95), na.rm=TRUE)
    hist(tmp, breaks=seq(0, max(a1, na.rm = TRUE), by=1), col="#808080", main = substr(colnames(a1)[i], 1, 7), xlab = "Read Depth")
    abline(v=sums[,i], col=2, lwd=2)
  
}
}


```

Now we can look at the per individual read depth for the most abundant allele for this Chilodonella vcf file.

```{r}

plot_read_depth(chilo_90)

```

